Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-16 iso-int/rc-iso-int.tex
small fixes
man ixes

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz});
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\item Sum types may similarly be represented by one wire or using
  parallel wires with a {{+}} operator between them. When tracing the
  execution of two additive wires, a value can reside on only one of the two
  wires.
\begin{multicols}{4}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}
\end{center}
%% \end{multicols}
%% \begin{multicols}{2}
\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
\includegraphics{diagrams/thesis/sum-two-wires.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
}
\end{center}
\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
}
\end{center}
\end{multicols}

@@

\item {{fold}} and {{unfold}} are basically drawn as triangular boxes
  that map the a recursive type to its unfolding.

\item {{trace+}} is drawn as a looped circuit, where the traced value
  is shown as flowing backward. The semantics and interpretation of
  {{trace}} are dealt with in detail in previous work (see Sec. 4 of
  \cite{infeffects}).

@@

\subsection{Adder}

Although trivial, the previous example captures the fundamental steps of our
construction. For reference, these steps are:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the left
  hand side of each machine state; 
\item Expand the output to expose enough structure to distinguish the right
  hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and introduce
  a {{trace}} to iterate the construction.
\end{enumerate}

Let us apply these steps to the slightly more interesting example of an
adder:
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m,p = 0 | n + 1
% Machine states = <e, e, e>

%subcode{bnf} include main
% Start state = <n, n, 0>
% Stop State = <0, n, n>
\end{multicols}
%subcode{opsem} include main
% <n+1, p, m> |--> <n, p+1, m+1>

@@

A closer look however reveals that the machine defines a \emph{partial}
isomorphism: not all valid final states can be mapped to valid start
states. Indeed consider the configuration {{<0,2,3>}} which is a valid final
state. Going backwards, the transitions start as follows 
{{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the machine gets stuck 
at a state that is not a valid start state.  There are several ways to deal with this
situation. The first is to use a partial isomorphism that diverges on the stuck state.
The second is to add another
possible kind of start state {{<n,0,n>}} to make the isomorphism total. 
Assuming the seond solution, our construction now proceeds as
expected. After expanding the input to expose $n$ and the output to expose
both $p$ and $m$, we reach the following configuration:

@@

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

We have also connected the matching input and output wires. At this point, we
have exposed both start states and the stop state. 
% \begin{enumerate}
% \item We can hide the use the combinator {{omega : b <-> 0}}
%   expressible in {{langRevT}}. This combinator is basically an
%   infinite loop on all inputs and hence would allow us to entirely
%   discard one possible branch. In other words, we are explicitly
%   admitting that our machine is undefined on some inputs.
% \item We can admit that there are two possible start states to this
%   abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
%   artifact of the fact that backward execution encounters stuck
%   states. Dually, if there were stuck states for forward execution,
%   they would be reflected in the fact that there were multiple stop
%   states -- only one of which denotes a desirable or correct execution
%   of the machine.
% \end{enumerate}
We can slide the two sides of the diagram and tie the knot using {{trace}} to
get the desired combinator:

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

@@

%%%%%%%%%%%%%%%%%%
\subsection{Tree Traversal}

The type of binary trees we use is {{@@x.nat+x*x}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are shown on the left:

\begin{multicols}{2}
%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = Lft c t | Rgt t c | []
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

%subcode{opsem} include main
% <L n, c> &|-->& {[c, L (n+1)]}
% <N t1 t2, c> &|-->& <t1, Lft c t2>
% {[Lft c t2, t1]} &|-->& <t2, Rgt t1 c>
% {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}

\end{multicols}

@@

%subcode{bnf} include main
%! columnStyle = rrcll
% unfold_t :& ~~~~t &<->& n + t * t &: fold_t
% unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c

\paragraph*{New Notation.}
To make the diagrams easier to understand, we introduce a syntactic
convenience which combines the first steps of the construction that consist
of {{fold}} / {{unfold}} and {{distribute}} / {{factor}}. We collectively
represent these steps using thin vertical rectangle. Also we will introduce
the convention that the component that is being expanded or collapsed will be
marked by using {{[^ ^]}} and the components that are being generated will be
marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
diagram below shows how to first expand the {{tree}} component and then in
one of the generated branches, expand the {{nat}} component:

@@

We can now apply our construction. The first step to developing the
isomorphic interpreter is to recognize that the two possible kinds of machine
states simply hide an implicit {{bool}}. We make this explicit:

\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}

@@

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{F}} machine state. For the sake of simplicity in
  this interpreter we won't be concerned with exposing stuck states on
  reverse execution caused by {{sub1 0}}. So we are free to use the {{add1}}
  combinator.

@@

This essentially completes the construction of the (partially)
isomorphic tree-traversal interpreter. (Partially because of the
implicit non-termination introduced by the {{add1}}).

%%%%%%%%%%%%%%%%%%%%%
\subsection{A {{langRevT}} Interpreter}

@@

%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14

@@

>>>>

In this paper, we investigate the practicality of writing recursive programs
in {{langRevT}}. Specifically, we assume that we are giving some reversible
recursive algorithm expressed as a \emph{reversible abstract machine}, and we
show via a number of systematic steps, how to develop a {{langRevT}} program
from that abstract machine. Our choice of starting from reversible abstract
machines is supported by the following observations:
\begin{itemize}
\item it is an interesting enough class of reversible programs: researchers
  have invested the effort in manually designing such machines, e.g., the
  SE(M)CD machine~\cite{Kluge:1999:SEMCD}, and the SECD-H~\cite{lorenz};
\item every reversible interpreter can be realized as such a machine: this
  means that our class of programs includes self-interpreters which are
  arguably a measure of the strength of any reversible
  language~\cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176,Yokoyama:2007:RPL:1244381.1244404};
\item general recursive programs can be systematically transformed to
  abstract machines: the technique is independent of reversible programs and
  consists of tranforming general recursion to tail recursion and then
  applying \emph{fission} to split the program into a driver and a small-step
  machine~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.
\end{itemize}

@@

\item Sum types may similarly be represented by one wire or using
  parallel wires with a {{+}} operator between them. When tracing the
  execution of two additive wires, a value can reside on only one of the two
  wires.
\begin{multicols}{4}
%\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-one-wire.pdf]
\includegraphics{diagrams/thesis/sum-one-wire.pdf}
}

%\end{center}
%% \end{multicols}
%% \begin{multicols}{2}
%\begin{center}
\scalebox{0.95}{
%%subcode-line{pdfimage}[diagrams/thesis/sum-of-wires.pdf]
\includegraphics{diagrams/thesis/sum-two-wires.pdf}
}

%\end{center}
%\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-left-value.pdf}
}

%\end{center}
%\begin{center}
\scalebox{0.95}{
\includegraphics{diagrams/thesis/sum-two-wires-right-value.pdf}
}
%\end{center}
\end{multicols}

@@

\item Operations {{fold}} and {{unfold}} are specific to the recursive
  data type and are drawn as triangular boxes. For instance,
  Sec. \ref{sec:bounded} shows {{fold}}/{{unfold}} isomorphisms for
  numbers, {{nat=@@x.(1+x)}}.

\item The {{trace+}} operation is drawn as a looped circuit, where the
  traced type {{b1}} is shown as flowing backward. The semantics and
  interpretation of {{trace}} are dealt with in detail in previous
  work (see Sec. 4 of \cite{James:2012:IE:2103656.2103667}).

@@

\subsection{Adder} 
\label{sec:adder}

Although trivial, the previous example captures the fundamental steps of our
construction. For reference, these steps are:
\begin{enumerate}
\item Expand the input to expose enough structure to distinguish the left
  hand side of each machine state; 
\item Expand the output to expose enough structure to distinguish the right
  hand side of each machine state;
\item Shuffle matching input terms to output terms, inserting any appropriate
  mediating computations. 
\item Slide the two sections to expose the start and stop state and
  introduce a {{trace}} to iterate the construction. This is always a
  mechanical process completely independant of the logic of the machine.
\end{enumerate}

Let us apply these steps to the slightly more interesting example of an
adder:

\vspace{-20pt}
\begin{small}
\begin{multicols}{2}  
%subcode{bnf} include main
% Numbers, n, m,p = 0 | n + 1
% Machine states = <e, e, e>

%subcode{bnf} include main
% Start state = <n, n, 0>
% Stop State = <0, n, n>
\end{multicols}
\vspace{-15pt}
%subcode{opsem} include main
% <n+1, p, m> |--> <n, p+1, m+1>
\end{small}

@@

A closer look however reveals that the machine defines a
\emph{partial} isomorphism: not all valid final states can be mapped
to valid start states. Indeed consider the configuration {{<0,2,3>}}
which is a valid final state. Going backwards, the transitions start
as follows {{<0,2,3> |--> <1,1,2> |--> <2,0,1>}} at which point, the
machine gets stuck at a state that is not a valid start state.

\paragraph*{Stuck States.}
All abstract machines, not just those written in {{langRevT}}, have to
deal with stuck states. There are several standard ways of dealing
with them:
\begin{itemize}
\item \emph{Use better types.}  One could augment {{langRevT}} with a
  richer type system that distinguishes non-zero numbers from those
  that can be zero, thereby eliminating the {{sub1 0}} situation
  entirely. Similarly, in the meta-circular intepreter in
  Sec. \ref{sec:langrevt-int} one could use GADTs to eliminate stuck
  states.

\item \emph{Divergence.}  Another standard approach in dealing with
  stuck states is to make the machine diverge or leave the output
  undefined/unobservable in some way. The {{add1}} combinator defined
  in Sec. 4.2 of our related work \cite{James:2012:IE:2103656.2103667}
  has the property that its dual {{sub1}} diverges on input 0. Hence
  we could just use {{add1}} here.

\item \emph{Stop the machine.}  If we chose to stop the machine at a
  stuck state, we are essentially treating states of the form
  {{<n,0,n>}} as valid stop states for reverse execution. This also
  has the consequence of making our isomorphism total and also gives
  us two valid start states in the case of forward execution. If we
  choose this approach, the machine would look as shown below by the
  end of {{step 3}}:

%% In the case of our machine this corresponds to another possible kind
%% of start state {{<n,0,n>}} to make the isomorphism total.  Let us
%% choose this path in the case of this machine.  After expanding the
%% input to expose $n$ and the output to expose both $p$ and $m$, we wire
%% the machine corresponding to the reduction rule.

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/adder3.pdf}
}
\end{center}

Note that there are indeed two ``start states'' in the interpreter. As before, we can
slide the two sides of the diagram and tie the knot using {{trace}} to
get the desired interpreter:

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/adder4.pdf}
}
\end{center}

More generally, how the programmer handles stuck states depends on the
specific program. For some machines it may be feasible to use a
stronger type system while for others, such as our tree example in
Sec. \ref{sec:tree}, it is acceptable to diverge.

%% We have also connected the matching input and output wires. At this
%% point, we have exposed both start states and the stop state.

%% There are several ways to deal with this situation. The first is to
%% use a partial isomorphism that diverges on the stuck state.

@@

% \begin{enumerate}
% \item We can hide the use the combinator {{omega : b <-> 0}}
%   expressible in {{langRevT}}. This combinator is basically an
%   infinite loop on all inputs and hence would allow us to entirely
%   discard one possible branch. In other words, we are explicitly
%   admitting that our machine is undefined on some inputs.
% \item We can admit that there are two possible start states to this
%   abstract machine {{<n, m, 0>}} and {{<n,0,m>}}. One of this is an
%   artifact of the fact that backward execution encounters stuck
%   states. Dually, if there were stuck states for forward execution,
%   they would be reflected in the fact that there were multiple stop
%   states -- only one of which denotes a desirable or correct execution
%   of the machine.
% \end{enumerate}

@@

%%%%%%%%%%%%%%%%%%
\subsection{Tree Traversal}
\label{sec:tree}

The type of binary trees we use is {{@@x.(nat+x*x)}}, i.e., binary trees with
no information at the nodes and with natural numbers at the leaves. To define
the abstract machine, we need a notion of \emph{tree contexts} to track which
subtree is currently being explored. The definitions are shown on the left:

\vspace{-15pt}
\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Tree, t = L n | N t t
% Tree Contexts, c = [] | Lft c t | Rgt t c
%
% Machine states = <t, c> | {[c, t]}
% Start state = <t, []>
% Stop State = {[[], t]}

%subcode{opsem} include main
% <L n, c> &|-->& {[c, L (n+1)]}
% <N t1 t2, c> &|-->& <t1, Lft c t2>
% {[Lft c t2, t1]} &|-->& <t2, Rgt t1 c>
% {[Rgt t1 c, t2]} &|-->& {[N t1 t2, c]}
\end{multicols}
\end{small}

@@

\begin{small}

%subcode{bnf} include main
%! columnStyle = rrcll
% unfold_t :& ~~~~t &<->& n + t * t &: fold_t
% unfold_c :& c &<->& 1 + c * t + t * c ~~~~~~&: fold_c
\end{small}

\paragraph*{New Notation.}
To make the diagrams easier to understand, we introduce a syntactic
convenience which combines the first steps of the construction that consist
of {{fold}} / {{unfold}} and {{distribute}} / {{factor}}. We collectively
represent these steps using thin vertical rectangle. Also we will introduce
the convention that the component that is being expanded (or constructed) will be
marked by using {{[^ ^]}} and the components that are being generated (or consumed) will be
marked by {{[v v]}}. For example, given a value of type {{tree*nat}}, the
diagram below shows how to first expand the {{tree}} component and then in
one of the generated branches, expand the {{nat}} component:

@@

We can now apply our construction. The first step to developing the
isomorphic interpreter is to recognize that the two possible kinds of machine
states simply hide an implicit {{bool=T~|~F}}. We make this explicit:

\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}
\end{small}

@@

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{F}} machine state. For the sake of simplicity, in
  this interpreter we won't be concerned with exposing stuck states on
  reverse execution caused by {{sub1 0}}. So we are free to use the {{add1}}
  combinator.

@@

This essentially completes the construction of the (partially)
isomorphic tree-traversal interpreter, except for the boring final
step. (Partially because of the implicit non-termination introduced by
the {{add1}}).

%%%%%%%%%%%%%%%%%%%%%
\subsection{A {{langRevT}} Interpreter}
\label{sec:langrevt-int}

In the final construction we present a meta-circular interpreter for
{{langRevT}} written in {{langRevT}}. This is a a non-trivial abstract
machine with several cases, but the deriving the corresponding
{{langRevT}} interpreter follows in exactly the same way as
before. Here is a logically reversible small-step abstract machine for
{{langRevT}} and the derived isomorphic interpreter with the
reductions labelled.

\begin{scriptsize}

@@

\vspace{-15pt}
%subcode{opsem} include main
%! columnStyle = rclr
% <iso, v, cc> &|-->& {[iso, iso(v), cc]} &~~~~~~~~~~~~ rule 1
% <c1(;)c2, v, cc> &|-->& <c1, v, Fst cc c2> &~~~~~~~~~~~~ rule 2
% {[c1, v, Fst cc c2]} &|-->& <c2, v, Snd c1 cc> &~~~~~~~~~~~~ rule 3
% {[c2, v, Snd c1 cc]} &|-->& {[ c1(;)c2, v, cc ]} &~~~~~~~~~~~~ rule 4
% <c1(+)c2, L v, cc> &|-->& <c1, v, LeftPlus cc c2> &~~~~~~~~~~~~ rule 5
% {[ c1, v, LeftPlus cc c2 ]} &|-->& {[c1 (+) c2, L v, cc ]} &~~~~~~~~~~~~ rule 6
% <c1(+)c2, R v, cc> &|-->& <c2, v, RightPlus c1 cc> &~~~~~~~~~~~~ rule 7
% {[ c2, v, RightPlus c1 cc ]} &|-->& {[c1 (+) c2, R v, cc ]} &~~~~~~~~~~~~ rule 8
% <c1(*)c2, (v1, v2), cc> &|-->& <c1, v1, LeftTimes cc c2 v2> &~~~~~~~~~~~~ rule 9
% {[ c1, v1, LeftTimes cc c2 v2 ]} &|-->& <c2, v2, RightTimes c1 v1 cc> &~~~~~~~~~~~~ rule 10
% {[ c2, v2, RightTimes c1 v1 cc ]} &|-->& {[ c1 (*) c2, (v1, v2), cc ]} &~~~~~~~~~~~~ rule 11
% < trace c, v, cc> &|-->& <c, R v, Trace cc> &~~~~~~~~~~~~ rule 12
% {[c, L v, Trace cc]} &|-->& <c, L v, Trace cc> &~~~~~~~~~~~~ rule 13
% {[c, R v, Trace cc]} &|-->& {[trace c, R v, cc]} &~~~~~~~~~~~~ rule 14

\end{scriptsize}

%% The isomorphic interpreter thus derived is shown below, just before
%% the `flipping' step (which makes it harder to read). Each wire has
%% been annotated with the reduction rule it corresponds to making it
%% easier to compare with the abstract machine.

The derivation of the {{langRevT}} interpreter, while tedious, is
entirely straightforward.  A couple of points however need clarification:

\begin{enumerate}

\item The definition here shows only the handling of the composition
  combinators, which are the interesting cases. All the primitive
  isomorphisms are hidden in the `{{iso(v)}}' application in
  {{rule~1}}. This should be read as `tranform {{v}} according to the
  primitive isomorphism {{iso}}'. Implementing this is straightformard
  and is hence skipped.

\item Stuck states in this machine correspond to runtime values not
  matching their expected types. They can be handled as decribed in
  Sec. \ref{sec:adder} using GADTs (which eliminates them entirely),
  divergence, or adding extra halting states. We have abtracted from
  this choice and marked the relevant cases with combinators labelled
  `{{cast}}'.

@@

We have developed a technique for the systematic development of
{{langRevT}} programs from logically reversible small-step abstract
machines. Since techniques for developing small-step interpreters are
well developed this allows for the direct development of a large class
of {{langRevT}} programs. We have demonstarted the effectiveness of
this approach by deriving meta-circular interpreter for {{langRevT}}
in this style.
